Semantic Linearity in a Lazy Optimising Compiler

### 2025-10-07 popl26-accepted-before-rebuttal.tex

    (3) $Let~\D',\D'' = \D,\Theta{\uparrow}^1$\\
    (4) $\G, \Theta{\uparrow}^\omega; \D' \vdash e : \s$\`by inv on let (2)\\
    (5) $\G, \Theta{\uparrow}^\omega; \D', \D'' \vdash (x, \Theta{\downarrow}^1_e, \Sigma) : \s \bigotimes(\Theta{\downarrow}^1_\s) \bigotimes(\Sigma)$\`by inv on let (2)\\
    (6) $\G, \Theta{\uparrow}^\omega; \D' \vdash x : \s$\\
    (7) $\G, \Theta{\uparrow}^\omega; \D'' \vdash (\Theta{\downarrow}^1_e, \Sigma) : \bigotimes(\Theta{\downarrow}^1_\s) \bigotimes(\Sigma)$\\
    (8) $\G, \Theta{\uparrow}^\omega; \D', \D'' \vdash (e, \Theta{\downarrow}^1_e, \Sigma) : \tau \bigotimes(\Theta{\downarrow}^1_\s) \bigotimes(\Sigma)$\` by pair (4,7)\\
    (9) $\G; \D \vdash \Theta \mid e : \tau,\Sigma$\` by well-typed state\\
    (10) $\G; \D \vdash \Theta \mid v : \tau,\Sigma$\` by induction hypothesis\\
    (11) $Let~\G';\D_a,\D_b = \G, \Theta'{\uparrow}^\omega, \D, \Theta'{\uparrow}^1$\\
    (11a) $\G'; \D_a \vdash v : \s$
    (11b) $\G'; \D_b \vdash (\Theta'{\downarrow}^1_e, \Sigma) : \bigotimes(\Theta'{\downarrow}^1_\s) \bigotimes(\Sigma)$\\
    (12) $\G'; \D_a \vdash \llet{x{:}_{\D_a}\s = v}{v} : \s$\`by let (11a,11a)\\
    (13) $\G; \D \vdash \Theta' \mid \llet{x{:}_{\D_a}\s = v}{v} : \s, \Sigma$\`by well-typed state\\
    (14) $\G; \D \vdash \Theta', x{:}_{\D_a}\s = v \mid v : \s, \Sigma$\`by env. expansion\\
\end{tabbing}
Note: It's slightly more elaborate to accommodate recursion. The runtime
environment bindings must be tagged with the recursive group such that the
environment expansion can decide whether to unwrap to a letrec group or
not. Unlike the case proved here, when the $\D$-binding belongs to a let group,
the whole group is unwrapped and we can invert the assumption to get an
$e$ typed with the $x$s from the group.\\
\item[Case:] $\Gamma; \Delta \vdash (\Theta \mid \ccase{e}{z{:}_{\Delta'}\s~\{K~\ov{y_i}\to e'}\}) \Downarrow (\Theta'' \mid v) : \tau$
\begin{tabbing}
    (1) $\Gamma; \Delta \vdash \Theta \mid \ccase{e}{z{:}_{\Delta_a}\s~\{K~\ov{y_i}\to e'}\} : \tau, \Sigma$\\
    (2) $\G, \Theta{\uparrow}^\omega; \D, \Theta{\uparrow}^1 \vdash (\ccase{e}{z{:}_{\Delta_a}\s~\{K~\ov{y_i}\to e'}\}, \Theta{\downarrow}^1_e, \Sigma) : \tau \otimes \bigotimes(\Theta{\downarrow}^1_\sigma) \otimes \bigotimes(\Sigma)$\`by \ref{lem:unwrapenv}\\
    (3) $Let~\G' = \G, \Theta\uparrow^\omega; \D_a,\D_b,\D_2,\D_3 = \D,\Theta\uparrow^1$\`s.t. $(\D_2,\D_3)$ types $(\Theta\downarrow^1_e, \Sigma)$\\
    (4) $\G'; \D_a, \D_b \vdash \ccase{e}{z{:}_{\D_a}\s\{K~\ov{y_i}\to e'\}} : \tau$\\
    Subcase $e$ is in WHNF (thus $= K~\ov{x_i}$)\\
    (5) $\G'; \D_a \Vdash K~\ov{x_i} : \s \gtrdot \ov{\D_i}$\` by subcase\\
    (6) $\G', z{:}_{\D_a}\s; \D_a, \D_b \vdash_{alt} K~\ov{y_i} \to e' :^z_{\D_i} \s \Rightarrow \tau$\` by inv\\
    Subsubcase $\vdash_{alt}$ $Alt_{WHNF}$\\
    (7) $\G', z{:}_{\D_a}\s, \ov{y_i{:}_\omega\s_i}, \ov{y_j{:}_{\Delta_j}\s_j}; \D_a, \D_b \vdash e' : \tau$\` by inv\\
    (8) $\Theta' = \Theta$\` in this subcase because $(\Theta \mid K~\ov{x_i})\Downarrow(\Theta \mid K~\ov{x_i})$\\
    (9) $\G'; \ov{\D_i} \vdash K~\ov{x_i} : \s$\` by inv (5)\\
    (10) $\ov{\G'; \cdot \vdash x_i : \s_i}$\` by inv (9) for unrestricted $x$s\\
    (11) $\ov{\G'; \D_j \vdash x_j : \s_j}$\` by inv (9) for linear fields $x$s\\
    (12) $\G', z{:}_{\D_a}\s; \D_a, \D_b \vdash e'\ov{[x_{i,j}/y_{i,j}] : \tau}$\` by $\D$-subst and $\omega$-subst lemmas (7,10,11)\\
    (13) $\G'; \D_a, \D_b \vdash e'\ov{[x_{i,j}/y_{i,j}]}[K~\ov{x_{i,j}}/z] : \tau$\` by $\D$-subst lemma and $\ov{\D_i} = \D_a$ (9,12)\\
    (14) $\G; \D \vdash \Theta \mid e'\ov{[x_{i,j}/y_{i,j}]}[K~\ov{x_{i,j}}/z] : \tau, \Sigma$\` by well-typed state\\
    (15) $\G; \D \vdash \Theta'' \mid v : \tau, \Sigma$\` by induction hypothesis\\
    Subsubcase $\vdash_{alt}$ $Alt_{0}$\\
    (7) $(\G', z{:}_{\D_a}\s)[\cdot/\D_a]_z, \ov{y_i{:}_\omega\sigma_i}; (\D_a, \D_b)[\cdot / \D_a] \vdash e' : \tau$\` by inv subcase\\
    (8) $\G', z{:}_{\cdot}, \ov{y_i{:}_\omega\sigma_i}; \D_b \vdash e' : \tau$\` by $z$ cannot occur in $\G'$\\
    (9) Since $\G';\D_a \Vdash K~\ov{x_i}$, in this subcase $\D_a = \cdot$\\
    (10) $\G', \z{:}_\cdot\sigma, \ov{y{:}_\omega\sigma}; \D_b \vdash e' : \tau$\` by (9)\\
    (11) $\G', \z{:}_\cdot\sigma; \D_b \vdash e'\ov{[x_i/y_i]} : \tau$\` by $\omega$-subst (10)\\
    (12) $\G'; \D_b \vdash e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] : \tau$\` by $\D$-subst\\
    (13) $\G; \D \vdash \Theta \mid e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] : \tau, \Sigma$\` by well-typed state\\
    (14) $\G; \D \vdash \Theta'' \mid v : \tau, \Sigma$\`by induction hypothesis\\
    Subcase $e$ is not in WHNF\\
    (5) $\G'; \D_a \vdash e : \s$\` by inv on case\\
    (6) $\G', z{:}_{\irr{\D_a}}\s; \irr{\D_a}, \D_b \vdash_{alt} K~\ov{y_{i,j}} \to e' :^z_{\irr{\D_a}} \s $\` by inv on case\\
    Subsubcase $\vdash_{alt}$ $Alt_{NWHNF}$\\
    (7) $\G', z{:}_{\irr{\D_a}}\s, \ov{y_i{:}_\omega\s_i}, \ov{y_j{:}_{\Delta_j}\s_j}; \irr{\D_a}, \D_b \vdash e' : \tau$\` for \ov{\D_j = \lctag{\irr{\D_a}}{K_j}}, by inv.\\
    (8) $\G', z{:}_{\ov{y_j{:}_1\s_j}}\s, \ov{y_i{:}_\omega\s_i}; \ov{y_j{:}_1\s_j}, \D_b \vdash e' : \tau$\` by $\D$-bound to linear\\
    (9) $\G', z{:}_{\ov{y_j{:}_1\s_j}}\s, \ov{y_i{:}_\omega\s_i}; \D_a, \D_b, \ov{y_j{:}_1\s_j} \vdash (e, e') : \sigma \otimes \tau$\` by pair (5,8)\\
    (10) $\G, z{:}_{\ov{y_j{:}_1\s_j}}\s, \ov{y_i{:}_\omega\s_i}; \D, \ov{y_j{:}_1\s_j} \vdash \Theta \mid e : \s, e':\tau, \Sigma$\`by well-typed state\\
    (11) $\G, z{:}_{\ov{y_j{:}_1\s_j}}\s, \ov{y_i{:}_\omega\s_i}; \D, \ov{y_j{:}_1\s_j} \vdash \Theta' \mid K~\ov{x_i} : \s, e':\tau, \Sigma$\`by induction hypothesis\\
    (12) $\G, \Theta'{\uparrow}^\omega, \ov{y_i{:}_\omega\s_i}; \D, \ov{y_j{:}_1\s_j}, \Theta'{\uparrow}^1 \vdash (K~\ov{x_{i,j}}, \Theta'{\downarrow}^1_e, e', \Sigma) : \s \bigotimes(\Theta'\downarrow^1_\sigma) \otimes \tau \bigotimes(\Sigma)$\`by \ref{lem:unwrapenv}\\
    (13) $Let~\G''=\G,\Theta'{\uparrow}^\omega~and~\D_c\D_d\D_4\D_5=\D,\Theta'{\uparrow}^1$\` where $(\D_4, \D_5)$ types $(\Theta'{\downarrow^1_e}, \Sigma)$\\
    (13a) $\G''; \D_c \vdash K~\ov{x_{i,j}} : \s$\\
    (13b) $\G'', z{:}_{\ov{y_j{:}_1\s_j}}\s, \ov{y_i{:}_\omega\s_i}; \D_d, \ov{y_j{:}_1\s_j} \vdash e' : \tau$\\
    (14) $\ov{\G''; \cdot \vdash x_i : \s_i}$\\
    (15) $\ov{\G''; \D_j \vdash x_j : \s_j}$\` for $\ov{D_j}=\D_c$, by inv\\
    (16) $\G'', z{:}_{\D_c}\s; \D_c, \D_d \vdash e'\ov{[x_{i,j}/y_{i,j}]} : \tau$\`by $1$-substs and $\omega$-substs lemmas\\
    (17) $\G''; \D_c, \D_d \vdash e'\ov{[x_{i,j}/y_{i,j}]}[K~\ov{x_{i,j}}/z] : \tau$\`by $\D$-subst lemma (13a,16)\\
    (18) $\G; \D \vdash \Theta' \mid e'\ov{[x_{i,j}/y_{i,j}]}[K~\ov{x_{i,j}}/z] : \tau, \Sigma$\`by well-typed state\\
    (19) $\G; \D \vdash \Theta'' \mid v : \tau, \Sigma$\`by induction hypothesis\\
    Subsubcase $\vdash_{alt}$ $Alt_0$\\
    (7) $\G', z{:}_\cdot\s, \ov{y_i{:}_\omega\s_i}; \D_b \vdash e' : \tau$\` by inv\\
    (8) $\G', z{:}_\cdot\s, \ov{y_i{:}_\omega\s_i}; \D_a, \D_b \vdash (e, e') : \sigma \otimes \tau$\` by pair (5,7)\\
    (9) $\G, z{:}_\cdot\s, \ov{y_i{:}_\omega\s_i}; \D \vdash \Theta \mid e : \sigma, e' : \tau, \Sigma$\` by well-typed state\\
    % (10) $\G, z{:}_\cdot\s, \ov{y_i{:}_\omega\s_i}; \D \vdash \Theta \mid e : \sigma, e' : \tau, \Sigma$\` by reconstructing\\
    (10) $\G, z{:}_\cdot\s, \ov{y_i{:}_\omega\s_i}; \D \vdash \Theta' \mid K~\ov{x_i} : \sigma, e' : \tau, \Sigma$\` by induction hypothesis\\
    (11) $Let~\G'' = \G,\Theta'{\uparrow}^\omega~and~\D_d,\D_4,\D_5 = \D, \Theta'{\uparrow}^1$\` where $(\D_4, \D_5)$ types $(\Theta'{\downarrow^1_e}, \Sigma)$\\
    (11a) $\G''; \cdot \vdash K~\ov{x_i} : \s$\`since $K$ has no linear fields in this subcase\\
    (11b) $\G'', z{:}_{\cdot}\s, \ov{y_i{:}_\omega\s_i}; \D_d \vdash e' : \tau$\\
    (12) $\ov{\G''; \cdot \vdash x_{i} : \s_i}$\`by inv (11a)\\
    (13) $\G'', z{:}_{\cdot}\s; \D_d \vdash e'\ov{[x_i/y_i]} : \tau$\` by $\omega$-subst lemma\\
    (14) $\G''; \D_d \vdash e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] : \tau$\` by $\D$-subst lemma\\
    (15) $\G; \D \vdash \Theta' \mid e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] :
    \tau, \Sigma$\` by well-typed state\\
    (16) $\G; \D \vdash \Theta'' \mid v : \tau, \Sigma$\` by induction hypothesis\\
\end{tabbing}
\item[Case:] $\Gamma; \Delta \vdash (\Theta \mid \ccase{e}{z{:}_{\Delta'}\s~\{\_ \to e'}\}) \Downarrow (\Theta'' \mid v) : \tau$
\begin{tabbing}
    (1) $\Gamma; \Delta \vdash \Theta \mid \ccase{e}{z{:}_{\Delta_a}\s~\{\_ \to e'}\} : \tau, \Sigma$\\
    (2) $\G, \Theta{\uparrow}^\omega; \D, \Theta{\uparrow}^1 \vdash (\ccase{e}{z{:}_{\Delta_a}\s~\{\_ \to e'}\}, \Theta{\downarrow}^1_e, \Sigma) : \tau \otimes \bigotimes(\Theta{\downarrow}^1_\sigma) \otimes \bigotimes(\Sigma)$\`by \ref{lem:unwrapenv}\\
    (3) $Let~\G' = \G, \Theta\uparrow^\omega; \D_a,\D_b,\D_2,\D_3 = \D,\Theta\uparrow^1$\`s.t. $(\D_2,\D_3)$ types $(\Theta\downarrow^1_e, \Sigma)$\\
    (4) $\G'; \D_a, \D_b \vdash \ccase{e}{z{:}_{\D_a}\s\{\_ \to e'\}} : \tau$\\
    (5) $\G'; \D_a \vdash e : \sigma$\\
    Subcase $e$ is in WHNF\\
    (-) Similar to the same subcase in the case-with-pattern-vars reduction,\\
        except only $Alt_{\_}$ is relevant which is far simpler\\
    Subcase $e$ is not in WHNF\\
    (6) $\G', z{:}_{\irr{\D_a}}\s; \irr{\D_a}, \D_b \vdash e' : \tau$\\
    (7) $\G'; \D_b, z{:}_1\s \vdash e' : \tau$\` by $\D$-bound to linear\\
    (8) $\G'; \D_a, \D_b, z{:}_1\s \vdash (e, e') : \sigma \otimes \tau$\` by pair\\
    (9) $\G; \D, z{:}_1\s \vdash \Theta \mid e : \sigma, e' : \tau,
    \Sigma$\` by well-typed state\\
    (10) $\G; \D, z{:}_1\s \vdash \Theta' \mid K~\ov{x_i} : \sigma, e' : \tau, \Sigma$\` by induction hypothesis\\
    (11) $Let~\G';\D_c,\D_d,\D_4,\D_5 = \G, \Theta'{\uparrow}^\omega; \Delta, \Theta'{\uparrow}^1$\`s.t. $(\D_4,\D_5)$ types $(\Theta'{\downarrow}^1_e, \Sigma)$\\
    (12) $\G';\D_c \vdash K~\ov{x_i} : \s$\\
    (13) $\G';\D_d,z{:}_1\s \vdash e' : \tau$\\
    (14) $\G';\D_c, \D_d \vdash e'[K~\ov{x_i}/z] : \tau$\` by $1$-subst lemma\\
    (15) $\G;\D \vdash \Theta' \mid e'[K~\ov{x_i}/z] : \tau, \Sigma$\` by well-typed state\\
    (16) $\G;\D \vdash \Theta'' \mid v : \tau, \Sigma$\` by induction hypothesis\\
\end{tabbing}
\end{description}
\end{proof}

We also derive type preservation for the lazy natural semantics.

  \begin{theorem}[Type Preservation]
For any well-typed $[\Theta \mid e]$, if $\Theta : e \Downarrow
\Theta' : e'$ or $\Theta : e \Downarrow^* \Theta' : e'$ then we have
that $[\Theta \mid e']$ is well-typed.
\end{theorem}
\begin{proof}
By Lemma~\ref{lem:naturaltoinstr} and Theorem~\ref{thm:typepres}.
\end{proof}

\subsection{Progress\label{sec:proof:progress}}

%\input{language-v4/proofs/ProgressTheorem}

\begin{theorem}[Progress]\label{thm:prog}
Let $\G ; \D \vdash \Theta \mid e : \tau,\Sigma$. Any partial derivation $\G; \D \vdash (\Theta \mid e)
\Downarrow?,\Sigma$ can be extended. 
\end{theorem}

\begin{proof}
 Essentially all cases are standard. The only potential issue is the
 (linear) variable rule.
It suffices to show that whenever a variable is evaluated it is
present in the evaluation environment. This situation is impossible
since the state
$\G ; \D \vdash \Theta \mid x : \tau , \Sigma$ where $x$ is absent
from $\Theta$ is not well-typed and so cannot be the head of a partial
derivation. 
\end{proof}

### 2025-10-07 popl26-accepted-before-rebuttal.tex

\begin{theorem}[Progress]
For any partial derivation of $\Theta : e \Downarrow?$, for $[\Theta |
e]$ well-typed, the derivation can be extended.
\end{theorem}
\begin{proof}
Follows from Lemma~\ref{lem:instrtonatural} and Theorem~\ref{thm:prog}.
 \end{proof}

\section{Optimisations preserve linearity}\label{app:optimisations}

\subsection{Inlining}
\input{language-v4/proofs/optimizations/Inlining}

\subsection{\texorpdfstring{$\beta$}{Beta}-reduction}

\input{language-v4/proofs/optimizations/BetaReduction}

\subsection{Case of known constructor}

\input{language-v4/proofs/optimizations/CaseOfKnownConstructor}

\subsection{Case of Case\label{sec:proof:caseofcase}}

The case of case transformation applies to case expressions whose scrutinee is
another case expression, and returns the innermost case expression transformed by repeating
the outermost case expression in each alternative of the innermost case,
scrutinizing the original alternative body.

Intuitively, since the scrutinee of the outermost case is not in WHNF, no
resources from it can directly occur in the outermost alternatives. By moving
the outermost alternatives inwards with a different scrutinee, the alternatives
remain well-typed because they are typed using either the case binder or the
pattern bound variables, which, by the \emph{Irrelevance} lemma, makes it
well-typed for any scrutinee consuming arbitrary resources.

\input{language-v4/proofs/optimizations/CaseOfCase}

\subsection{Commuting let}

\input{language-v4/proofs/optimizations/LetFloating}

\subsection{\texorpdfstring{$\eta$}{Eta}-conversions}

\input{language-v4/proofs/optimizations/EtaConvs}

\subsection{Binder Swap}

The binder swap transformation applies to case expressions whose scrutinee is a
single variable $x$, and it substitutes occurrences of $x$ in the case
alternatives for the case binder $z$. If $x$ is a linear resource, $x$ cannot
occur in the case alternatives (as we conservatively consider variables are not
in WHNF), so the substitution preserves types vacuously. Otherwise, $x$ can be
freely substituted by $z$, since $z$ is also an unrestricted resource (it's
usage environment is empty because $x$ is unrestricted).

\input{language-v4/proofs/optimizations/BinderSwap}

% G ; D |- (Theta | e x) \Downarrow (Theta'' | v) : tau, Sigma

% QUEREMOS PROVAR:
% Se
% G ; D |- (Theta | e x) : tau , Sigma
% entao
% G ; D |- (Theta'' | v) : tau, Sigma

% por i.h.
% Se G;D |- (Theta | e) : sigma -> tau, x:tau , Sigma
% entao
% G ; D |- (Theta' | \y:sigma.e') : sigma->tau, x:tau,Sigma

% por i.h.
% Se G ; D |- (Theta' , y:sigma = x | e') : tau, Sigma
% entao
% G;D |- (Theta'' | v) : tau , Sigma

% Gamma ; Delta |- [Theta | (e x , terms(Sigma))] : (tau , types(Sigma))
% Gamma ; Delta |- [Theta | (e , terms(x:tau,Sigma))] : (tau->sigma , types(x:tau,Sigma))

% Daqui Gamma ; Delta |- [Theta' | (\y:_1 sigma.e' , terms(x:tau,Sigma))] : (sigma->tau, types(x:tau,Sigma))
% tirar Gamma ; Delta |- [Theta',y:sigma = x | (e' , terms(Sigma))] : (sigma->tau, types(Sigma))

% [Theta' , x_1 : tau = y | e] = (\x_1:tau . [Theta' | e] ) y
% [ Theta' , x_Delta:tau = e' | e] = let x_Delta = e' in [Theta' | e]
% [ . | (e,e')] = (e,e')

% Gamma ; Delta |- [Theta | (e , terms(Sigma))] : (tau , types(Sigma))

 % Gamma ; Delta |- Theta | e  : tau, Sigma

% }}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\end{document}

% TODO: - In the letrec case: congratulations on finding an inference algorithm
% (though I will want to see the proof some day, I don't see why it works yet).
% You should clarify that, during linting, you will have a usage environment
% annotation and won't need to run the inference algorithm. This algorithm is
% only needed when you first create a letrec.

% vim: fen fdm=marker
